---
title: "Data analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
rm(list = ls())
```

## Overview 

This script analyzes the relationship between PM2.5 and asthma rates in Colorado through various Poisson regression models. The main goal is to assess the impact of PM2.5 and access to health facilities on asthma rates, both in monthly and fire season contexts.

## Load packages

```{r}
library(tidyverse)
library(here)
library(sf)
library(tmap)
library(modelsummary)
library(patchwork)
```

## Load data

```{r}
# monthly PM2.5 & asthma data 
co_monthly_asthma_clean <- read_csv(here("data/new_data/PM2_5-asthma_counties/PM2_5_asthma_CO_monthly.csv"))

# PM2.5 & asthma data aggregated by fire season
co_fireszn_asthma_clean <- read_csv(here("data/new_data/PM2_5-asthma_counties/PM2_5_asthma_CO_fire_szn.csv"))
```

## Data exploration

Examine the distribution of asthma rates in Colorado. Given that asthma rates are essentially a count of the number of hospitalizations and ED visits, we expect the data to be right-skewed.

```{r}
# create a histogram to look at the distribution of the `rate` data 
rate_distribution <- ggplot(data = co_monthly_asthma_clean, aes(x = rate)) +
  geom_histogram(bins = 30, fill = "orange", color = "black") +
  labs(title = "Distribution of asthma rates in Colorado",
       x = "Asthma rate",
       y = "Count") + 
  theme_classic()
rate_distribution

# save the plot
ggsave(here("plots/asthma_rate_distribution.png"), 
       plot = rate_distribution, 
       width = 8, 
       height = 6)
```

Because the data is right-skewed and a count of events, we will use a Poisson regression model to analyze the relationship between PM2.5, health facilities, and asthma rates. 

## Data analysis

#### Processing 

Add a new variable of asthma rates lagged by 1 month. This will allow us to assess the impact of PM2.5 on asthma rates in the following month. 

```{r}
# add a new variable of asthma rates lagged by 1 month
co_monthly_asthma_clean <- co_monthly_asthma_clean %>%
  mutate(lag_rate = lag(rate, 1)) # lag rates by 1 month
```

Because our dependent variable is right-skewed, we log-transformed the data to normalize its distribution. 

```{r}
### standardize variables 

# monthly data
co_monthly_asthma_clean$log.rate <- log(co_monthly_asthma_clean$rate + 1)
co_monthly_asthma_clean$log.lag_rate <- log(co_monthly_asthma_clean$lag_rate + 1)
co_monthly_asthma_clean$log.smoke_pm_pred <- log(co_monthly_asthma_clean$smoke_pm_pred + 1)
co_monthly_asthma_clean$log.health_facility_count <- log(co_monthly_asthma_clean$health_facility_count + 1)

# fire season
co_fireszn_asthma_clean$log.rate <- log(co_fireszn_asthma_clean$rate + 1)
co_fireszn_asthma_clean$log.smoke_pm_pred <- log(co_fireszn_asthma_clean$smoke_pm_pred + 1)
co_fireszn_asthma_clean$log.health_facility_count <- log(co_fireszn_asthma_clean$health_facility_count + 1)
```

#### Poisson regression

Generate various models with asthma rates as the response variable, and PM2.5, health facility count, and year as predictors.

###### 2011-2023 (using complete monthly data)

Run two regressions with monthly asthma rates as the response variable: **(1)** current monthly observations of asthma rates and PM2.5 predictions; **(2)** lagged asthma rates by one month with current PM2.5 predictions. 

```{r}
# log asthma rates by log PM2.5 and health facility count
model_pois1 <- fixest::fepois(fml = log.rate ~ log.smoke_pm_pred + 
                                  log.health_facility_count | year, # year as FE
                                data = co_monthly_asthma_clean)
summary(model_pois1)

# log lagged asthma rates by log PM2.5 and health facility counts
model_pois2 <- fixest::fepois(fml = log.lag_rate ~ log.smoke_pm_pred + 
                                  log.health_facility_count | year, # year as FE
                                data = co_monthly_asthma_clean)
summary(model_pois2)
```

###### Fire seasons 2011-2023 (aggregating monthly data to fire season)

Run a regression with fire season asthma rates as the response variable, and fire season PM2.5 predictions and health facility count as predictors. 

```{r}
# log asthma rates by log PM2.5 and health facility count
model_pois3 <- fixest::fepois(fml = log.rate ~ log.smoke_pm_pred + 
                                  log.health_facility_count | year, 
                                data = co_fireszn_asthma_clean)
summary(model_pois3)
```

## Plotting the results 

#### Make a table combining all model outputs

```{r}
### make a table of the results

# make a list of all models
model_list <- list("Model 1 (monthly asthma & PM2.5)" = model_pois1, 
                   "Model 2 (lagged monthly asthma & PM2.5)" = model_pois2,
                   "Model 3 (fire season asthma & PM2.5)" = model_pois3)

# export summary table as a word doc
model_summary <- modelsummary::modelsummary(model_list, 
                           output = here("plots/tables/model_results.docx"), 
                           gof_omit = "IC|Log|Pseudo",  # Omits unnecessary goodness-of-fit stats
                           stars = TRUE)

# export summary table as a png
modelsummary::modelsummary(model_list, 
                           output = here("plots/tables/model_results.png"), 
                           gof_omit = "IC|Log|Pseudo",  # Omits unnecessary goodness-of-fit stats
                           stars = TRUE)
```

#### Make a plot of the results

```{r}
### make some plots of the result coefficients

# model_pois1
plot1 <- sjPlot::plot_model(model_pois1, 
                            type = "est",
                            sort.est = TRUE, 
                            show.values = TRUE, 
                            value.offset = 0.3, 
                            value.size = 2.5,
                            transform = NULL) + 
  labs(title = "A",
       x = " ",
       y = " ") + 
  theme_bw()

# model_pois2
plot2 <- sjPlot::plot_model(model_pois2, 
                            type = "est",
                            sort.est = TRUE, 
                            show.values = TRUE, 
                            value.offset = 0.3, 
                            value.size = 2.5,
                            transform = NULL) + 
  labs(title = "B",
       x = "Variable",
       y = " ") + 
  theme_bw()

# model_pois3
plot3 <- sjPlot::plot_model(model_pois3, 
                            type = "est",
                            sort.est = TRUE, 
                            show.values = TRUE, 
                            value.offset = 0.3, 
                            value.size = 2.5,
                            transform = NULL) + 
  labs(title = "C",
       x = " ",
       y = "Coefficient estimate") + 
  theme_bw()

# combine the plots with patchwork 
combined_plot <- plot1 + plot2 + plot3 +
  plot_layout(ncol = 1)
combined_plot
```

## Archive

#### Basic linear regression

```{r}
# # model asthma rates by PM2.5
# model <- lm(rate ~ smoke_pm_pred + factor(year) + factor(name), 
#             data = co_monthly_asthma_clean, 
#             na.action = na.exclude)
# 
# summary(model)
# 
# 
# # model log asthma rates by log PM2.5
# model_log <- lm(log.rate ~ log.smoke_pm_pred + factor(year) + factor(name), 
#                 data = co_monthly_asthma_clean, 
#                 na.action = na.exclude)
# summary(model_log)
# 
# 
# # model lagged asthma rates by PM2.5
# model_lag <- lm(lag_rate ~ smoke_pm_pred + factor(year) + factor(name), 
#                 data = co_monthly_asthma_clean, 
#                 na.action = na.exclude)
# summary(model_lag)
# 
# 
# # model lagged log asthma rates by log PM2.5
# model_lag_log <- lm(log.lag_rate ~ log.smoke_pm_pred + factor(year) + factor(name), 
#                     data = co_monthly_asthma_clean, 
#                     na.action = na.exclude)
```












